---
title: "PSRF Data Workshop 2026"
author: "Marina Kochuten"
format: html
editor_options: 
  chunk_output_type: console
---

# Intro to Quarto & Markdown

::: callout-tip
Quarto was Posit's follow up to R markdown files.

Quarto documents are fully reproducible and support dozens of output formats, like PDFs, Word files, presentations, and more (my website is built with Quarto!).

Quarto files are designed to be used in three ways:

For communicating to decision-makers, who want to focus on the conclusions, not the code behind the analysis.

For collaborating with other data scientists (including future you!), who are interested in both your conclusions, and how you reached them (i.e. the code).

As an environment in which to do data science, as a modern-day lab notebook where you can capture not only what you did, but also what you were thinking.
:::

## Writing Markdown Text

# This is a level one header

## This is a level two header

### This is a level three header

#### And so on

**This text is bold**

*This text is italic*

And this:

-   is
-   a
-   list

Here, I can link a [super cool website](https://restorationfund.org)

**To write code** 

Create a code chunk with the keyboard shortcut `Ctrl+Alt+I` (or `Cmd+Option+I` on Mac)

To run it, use `Ctrl+Shift+Enter` (Mac users: `Cmd+Shift+Enter`)

```{r}
2 + 2
```

If we were to render this into a document, you would see the code as well as the output. We can change this by using code-chunk options:

```{r}
#| code-fold: true

# This code will only show if you click on a dropdown to open it
2 + 2
```

```{r}
#| echo: false

# This code will not show at all, only the output
2 + 2
```

```{r}
#| include: false

# This code will run, but the code nor the output will show up in the rendered document
2 + 2
```


::: callout-tip
This is just a very brief intro to using Markdown in Quarto.

To learn more about Markdown, inserting tables and figures, and more see this [Comprehensive Guide to Using Quarto](https://quarto.org/docs/guide/)
:::


# Penguiuns Analysis

## About

This script hosts the code to clean and visualize data about the Palmer Penguins for the purpose of teaching some data wrangling best practices to the amazing PSRF team! 

**About the data**

There are 2 data files used in this analysis:

- `messy_penguins.csv`: contains size measurements for three penguin species observed on three islands in the Palmer Archipelago, Antarctica
- `nesting.csv`: contains nesting site counts for each species, on each island

`messy_penguins.csv` is derived from the Palmer Penguins data set from the Environmental Data Initiative (EDI) Data Portal. These data were collected from 2007 - 2009 by Dr. Kristen Gorman with the Palmer Station Long Term Ecological Research Program, part of the US Long Term Ecological Research Network. This data set was purposefully "messied" in order to practice some important data wrangling techniques.

`nesting.csv` is entirely constructed from my imagination for teaching purposes!

## Setup

```{r}
# Load libraries ----
library(tidyverse)
library(here)
library(janitor)
library(lubridate)

# Read in data ----
penguins_raw <- read_csv(here("data", "raw", "messy_penguins.csv"))
nesting_raw <- read_csv(here("data", "raw", "nesting.csv")) 
```


## Data exploration

Always take some time to get to know the data. Understanding the underlying structure of your data ahead of time makes cleaning and analysis much easier later!

### Penguins data

```{r}
# What columns are in the penguins data?
colnames(penguins_raw)

# How many penguins species are in the data?
n_distinct(penguins_raw$Species)

# What are they?
unique(penguins_raw$Species)

# What islands?
unique(penguins_raw$island)
# Looks like we will need to clean up this column later
```

### Nesting

```{r}
# What columns are in the Island info data?
colnames(island_info_raw)
```

## Data Cleaning

### Penguins

```{r}
# Peek at the data
head(penguins_raw)
```


```{r}
# Clean up column names for consistency
penguins_clean <- penguins_raw |>
  clean_names()

# Check column name fix
colnames(penguins_clean)
```


```{r}
# Species column ----
unique(penguins_clean$species) # looks good!
```


```{r}
# Island column ----
# Fix inconsistent capitalization in island column
penguins_clean <- penguins_clean |>
  mutate(island = str_to_title(island))

# Check island column capitalization fix
unique(penguins_clean$island)
```

```{r}
# Bill length column ----
range(penguins_clean$bill_length_mm, na.rm = TRUE)  # hmm... that's not right. 

# Not only is -999 not a valid bill length, but I would expect this column to be strictly numeric
class(penguins_clean$bill_length_mm) # Character, when would expect numeric

# What values are not recognized as a numeric?
penguins_clean |>
  count(bill_length_mm) |>
  filter(is.na(as.numeric(bill_length_mm)))

# It is common to have multiple funky ways to encode NAs. Let's change that now
penguins_clean <- penguins_clean |>
  mutate(bill_length_mm = as.numeric(bill_length_mm)) 
# Using as.numeric will make anything that R doesn't know how to make a number NA
```

I want to note that this is one way to carefully handle encoding NAs. I will also frequently do this when I read in the data, especially if I already know how NAs are handled. You can do this using the `na =` argument in `read_csv()`.

```{r}
# Now we can make sure the range of values for bill length makes sense
range(penguins_clean$bill_length_mm, na.rm = TRUE)

# Still need to handle -999
penguins_clean <- penguins_clean |>
  mutate(bill_length_mm = case_when(bill_length_mm == -999.0 ~ NA,
                                    .default = bill_length_mm))

# Check
range(penguins_clean$bill_length_mm, na.rm = TRUE) # much better
```

In this case, I am going to skip over the next few columns and get to `observation_date`. I highly recommend taking the time to check through each column and look for anything unexpected at this stage in the data process! 

```{r}
# Observation date column ----
head(penguins_clean$observation_date) # yikes, what a mess

# Fix datetime column
penguins_clean <- penguins_clean |>
  mutate(date = parse_date_time(observation_date,
                                orders = c("dmy", "ymd", "mdy", "ymd HMS")))

head(penguins_clean$date)

# Do not need to keep the time
penguins_clean <- penguins_clean |>
  separate(col = date,
           into = c("date", "time"),
           sep = " ") |>
  # Need to re-convert to date after using separate()
  mutate(date = as.Date(date))

# Drop date columns we don't need
penguins_clean <- penguins_clean |>
  select(-c(time, observation_date))

# Create some new variables based on date
penguins_clean <- penguins_clean |>
  mutate(year = year(date),
         month = month(date, label = TRUE, abbr = FALSE),
         day = day(date),
         day_of_year = yday(date))

# Check out changes
head(penguins_clean)
```

Other good ideas when cleaning data is to check for duplicates

```{r}
# Check for duplicate rows
penguins_clean |>
  get_dupes()
```

Deciding how to handle duplicates can be tricky. Is it possible that two penguins measured on the same day have the exact same measurements? Perhaps! It is possible that there was an error made during data entry? Also likely! Things I like to think about before removing duplicates:

- Check field notes or datasheet scans
- Check with field team / whoever entered the data when possible
- Only a few duplicate observations in a large data set? Won't change much to remove
- Lot's of duplicates, removing would take out more than 5% - 10% of the data? Reconsider

Whatever you decide, the most important thing is that you document what you are doing and why. The beauty about making these changes 1. not in the raw data and 2. with code is that it is very easy to go back if you change your mind.

For this case, let's assume it was a data entry error and remove the duplicate rows from the dataset entirely.

```{r}
# Remove duplicates in penguins data ----
# Count rows before removing
nrow(penguins_clean)

# Remove duplicates
penguins_clean <- penguins_clean |>
  distinct()

# Count rows after removing
nrow(penguins_clean)
```


```{r}
# Save cleaned penguin data ----
write_csv(penguins_clean, here("data", "processed", "penguins.csv"))
```

Now that we have completed this chuck of data cleaning, let's push to GitHub. 

## Some exploratory viz




## Nesting Sites

Let's say we are interesting in comparing the number of female penguins counted and the number of nesting sites each year. Since these data live in different datasets we need to join them! 

First, we need to get our nesting data in tidy format:

```{r}
# Pivot nesting data from wide to long
nesting_clean <- nesting_raw |>
  pivot_longer(cols = starts_with("nesting"),
               names_to = "year",
               names_prefix = "nesting_sites_",
               values_to = "nesting_sites")
```

And, to make things easier later let's count female penguins now:

```{r}
# Count the number of female penguins for each species, island, and year
penguins_fem_count <- penguins_clean |>
  group_by(species, island, sex, year) |>
  summarise(count = n()) |>
  filter(sex == "female")
```


Next, we need to choose which join to use. There are many kinds, here are a few:

- **Full join**: Keep all observations from both datasets
- **Left join**: Keep all observations in dataset y that have matches in x
- **Anti join**: Keep only observations in x that have no matches in y

When you join two datasets, they need a shared key to join on.

In our case, we want to use a left join, joining the nesting data to the penguins on species, island, and year.

```{r}
# Join penguins and nesting data
pen_nest <- left_join(penguins_fem_count, nesting_clean,
                      by = c("species", "island", "year"))

# Error - need to make year in nesting_clean numeric
nesting_clean <- nesting_clean |>
  mutate(year = as.numeric(year))

# Now try joining penguins and nesting data
pen_nest <- left_join(penguins_fem_count, nesting_clean,
                      by = c("species", "island", "year"))
```

Now we can visualize the relationship!

```{r}
# Exploratory viz - female penguin count vs number of nesting sites
ggplot(pen_nest) +
  geom_point(aes(x = nesting_sites, y = count, color = species)) +
  labs(title = "Penguins counted vs number of nesting sites")
```


# Exploring viz to make

```{r}
# Penguin counts bar chart
counts_bar <- penguins_clean |>
  mutate(species = fct_rev(fct_infreq(species))) |>
  ggplot() +
  geom_bar(aes(x = species, fill = species)) +
  geom_text(aes(x = species, label = after_stat(count)),
            stat = "count",
            hjust = 1.5,
            color = "white",
            size = 5) +
 scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  labs(title = "Penguins counts at Palmer Station LTER",
       subtitle = "2007 - 2009",
       caption = "Data: Gorman et al. 2014 via {palmerpenguins}") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    panel.grid = element_blank(),
    plot.title.position = "plot",
    axis.text.x = element_blank(),
    plot.margin = margin(20,20,20,20),
    plot.caption = element_text(face = "italic",
                                color = "grey50"))

ggsave(here("figs", "counts_bar.png"), width = 6)
```

```{r}
# highlights and annotations ----
# ex 1:
ggplot(data = penguins_clean, aes(x = body_mass_g)) +
  geom_histogram(aes(fill = species), 
                 alpha = 0.5, 
                 position = "identity") +
  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  gghighlight(species == "Chinstrap", label_key = type) +
  labs(x = "Body mass (g)",
       y = "Frequency",
       title = "Penguin body mass",
       fill = "Species") +
  theme_minimal()

# ex 2:
ggplot(data = penguins_clean, aes(x = body_mass_g)) +
  geom_histogram(aes(fill = species), 
                 alpha = 0.5, 
                 position = "identity") +
  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  gghighlight() +
  facet_wrap(~species) +
  labs(x = "Body mass (g)",
       y = "Frequency",
       title = "Penguin body mass") +
  theme_minimal()

# ex 3:
ggplot(data = penguins,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 3,
             alpha = 0.8) +
  scale_color_manual(values = c("darkorange","purple","cyan4")) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "Penguin size, Palmer Station LTER",
       subtitle = "Flipper length and body mass for Adelie, Chinstrap and Gentoo Penguins",
       x = "Flipper length (mm)",
       y = "Body mass (g)",
       color = "Penguin species",
       shape = "Penguin species") + 
  ggforce::geom_mark_ellipse(aes(filter = species == "Gentoo", 
                                 label = "Gentoo penguins", 
                                 description = "This species tends to have larger body masses and flipper lengths than the other species"),
                             color = "cyan4") +
  theme_minimal(base_size = 16) +
  theme(legend.position = c(0.2, 0.8),
        legend.background = element_rect(fill = "white", color = "white"),
        plot.title.position = "plot",
        plot.margin = margin(10,10,10,10),
        axis.title.x = element_text(margin = margin(15, 0, 0, 0)),
        axis.title.y = element_text(margin = margin(0, 15, 0, 0)),
        plot.subtitle = element_text(margin = margin(0,0,15,0))
  )
```

```{r}
# legend options ----
# 1. move legend inside of plot to reduce eye movement
ggplot(data = penguins_clean,
       aes(x = bill_length_mm,
           y = bill_depth_mm,
           group = species)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 3,
             alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, aes(color = species)) +
  scale_color_manual(values = c("darkorange","purple","cyan4")) +
  labs(title = "Penguin bill dimensions",
       subtitle = "Bill length and depth for Adelie, Chinstrap and Gentoo Penguins at Palmer Station LTER",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       color = "Penguin species",
       shape = "Penguin species") +
  theme_minimal(base_size = 14) +
  theme(legend.position = c(0.85, 0.15),
        plot.title.position = "plot",
        plot.caption = element_text(hjust = 0, face= "italic"),
        plot.caption.position = "plot",
        plot.margin = margin(10,10,10,10),
        axis.title.x = element_text(margin = margin(15, 0, 0, 0)),
        axis.title.y = element_text(margin = margin(0, 15, 0, 0)),
        plot.subtitle = element_text(margin = margin(0,0,15,0)))


# 2. incorporate legend into the title
ggplot(data = penguins_clean,
       aes(x = bill_length_mm,
           y = bill_depth_mm,
           group = species)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 3,
             alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, aes(color = species)) +
  scale_color_manual(values = c("darkorange","purple","cyan4")) +
  labs(title = "Penguin bill dimensions",
       
       # Subtitle with HTML tags
       subtitle = "Bill length and depth for
       <span style='color:darkorange;'>**Adelie**</span>, 
       <span style = 'color:purple;'>**Chinstrap**</span>, 
       and <span style = 'color:cyan4;'>**Gentoo**</span>,
        penguins at Palmer Station LTER",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       color = "Penguin species",
       shape = "Penguin species") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",  # Remove the original legend
        plot.title.position = "plot",
        plot.subtitle = ggtext::element_markdown(), # To let R know you are styling the text
        plot.caption = element_text(hjust = 0, face= "italic"),
        plot.caption.position = "plot",
        plot.margin = margin(10,10,10,10),
        axis.title.x = element_text(margin = margin(15, 0, 0, 0)),
        axis.title.y = element_text(margin = margin(0, 15, 0, 0)))


# 3. Label the lines directly
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species, shape = species)) +
  geom_point(size = 3, alpha = 0.8) +
  geomtextpath::geom_labelsmooth(aes(label = species), method = "lm", size = 5) +
  scale_color_manual(values = c("darkorange","purple","cyan4")) +
  labs(title = "Penguin bill dimensions",
       subtitle = "Bill length and depth for Adelie, Chinstrap and Gentoo Penguins at Palmer Station LTER",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       color = "Penguin species",
       shape = "Penguin species") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",  # Remove the original legend
        plot.title.position = "plot",
        plot.caption = element_text(hjust = 0, face= "italic"),
        plot.caption.position = "plot",
        plot.margin = margin(10,10,10,10),
        axis.title.x = element_text(margin = margin(15, 0, 0, 0)),
        axis.title.y = element_text(margin = margin(0, 15, 0, 0)))
```



## Using some index data?

```{r}
# Read in data ----
temp <- read_rds(here("data", "raw", "logger_data_temp.rds"))
```

```{r}
# Let's get monthly high
temp_agg <- temp |>
  separate(col = datetime, into = c("date", "time"), sep = " ") |>
  mutate(date = as.Date(date),
         year = year(date),
         month = month(date)) |>
  group_by(site, date) |>
  summarise(daily_high = max(tidbit_temp_c)) |>
  ungroup()

ggplot(temp_agg, aes(x = date, y = daily_high, color = site)) +
  geom_line() +
  theme_minimal(base_size = 12)

ggplot(temp_agg, aes(x = site, y = daily_high, color = site)) +
  geom_jitter(alpha = 0.3) +
  geom_boxplot(fill = NA) +
  scale_color_brewer(type = "qual", palette = 2) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme()

ggplot(temp_agg, aes(x = daily_high, y = site, fill = daily_high)) +
  geom_density_ridges_gradient(alpha = 0.5) +
  labs(x = "Daily High Temperature",
       title = "Daily High Temperatures Across Puget Sound Monitoring Sites",
       caption = "Data: 2025, Puget Sound Restoration Fund") +
  scale_x_continuous(labels = scales::label_glue("{x}ºC")) +
  scale_fill_viridis_c() +
  theme_minimal(base_size = 14) +
  theme(axis.title.y = element_blank(),
        plot.title.position = "plot",
        plot.caption = element_text(margin = margin(15, 0, 0, 0),
                                    color = "grey",
                                    face = "italic"),
        plot.caption.position = "plot",
        legend.position = "none",
        axis.title.x = element_text(margin = margin(15, 0, 0, 0)))

temp_agg$site <- factor(temp_agg$site)

ggplot(temp_agg, aes(x = daily_high, y = site, fill = ..x..)) +
  scale_fill_viridis_c(option = "C") +
  ggridges::geom_density_ridges_gradient(aes( alpha = 0.2)) +
  labs(
    x = "Daily High Temperature",
    title = "Daily High Temperatures Across Puget Sound Monitoring Sites",
    caption = "Data: 2025, Puget Sound Restoration Fund"
  ) +
  scale_x_continuous(labels = scales::label_glue("{x}ºC")) +
  
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    plot.title.position = "plot",
    plot.caption = element_text(margin = margin(15, 0, 0, 0),
                                color = "grey",
                                face = "italic"),
    plot.caption.position = "plot",
    legend.position = "none",
    axis.title.x = element_text(margin = margin(15, 0, 0, 0))
  )


```
